Towards a Complete Model of the Negative Selection Process in Humans

### 2023-06-13 manuscript/manuscript.tex
new ms stuff

Overall, we find the PhastCons Priority and CADD 6\% models fit about equally
as well (Figure \ref{fig:figure-2}A), explaining $R^2=68.45$\% and
$R^2=68.04$\% of the out-sample variance in Yoruba pairwise diversity at the
megabase-scale, respectively. In populations impacted by the out-of-Africa
bottleneck, the goodness-of-fit was lower (62.98\% and 59.98\% for CEU and CHB,
respectively). The CADD 6\% model was previously found by
\textcite{Murphy2022-sj} to be the best-fitting model using classic BGS. Our
PhastCons Priority model fits marginally better than the CADD 6\% model
consistently across all three populations (Supplementary Table XXX). This is
likely because the PhastCons Priority model has more DFE parameters, and thus
can fit finer-scale differences in the DFE across its four feature types. By
contrast, the CADD 6\% track must fit patterns of diversity to conserved sites
under a single DFE. The CADD 8\% model fit slightly worse than the CADD 6\%
across all populations, and the Feature Priority model had the poorest-fit.

Since our method is built upon theory that fixes the weak selection problem of
classic BGS theory, it should in principle fit equally well when the annotation
model covers the full genome, the bulk of which is expected to be neutrally
evolution. To test this, for each annotation model (which is ``sparse") we fit
a ``full" model that contains the remaining complement of the genome labeled as
``other". Ideally, the goodness-of-fit of fit using our B' method should be
invariant to whether an annotation model is sparse or full. Indeed, we find
that

NOTE: there looks like a bug in the table... repeated values.

Each model was fit to the high-coverage 1000 human genome data for
three populations, YRI, CEU, and CHB. Additionally, we also fit each of these
models using our method, but using classic BGS theory rather than our B'
theory. In total, we fit and compared 48 different models.

Overall, we find that the goodness-of-fit, measured by both in- and out-sample
$R^2$ are nearly identical across sparse and full fits using our B' method,
regardless of the different annotation models (Figure \ref{fig:figure-2},
Supplementary Materials Table XXX). By contrast, we find that when set our
method to use classic BGS theory, the model fits are worse in full-coverage
annotation models (Figure \ref{fig:figure-2}B). Additionally, when sparse
annotation models contain genomic features that are likely under weak
constraint (such as introns and UTRs), models fit worse under classic BGS
theory than our B' method (Figure \ref{fig:figure-2}A). However, among the CADD
annotation models, the goodness-of-fit is nearly identical between B' and
classic BGS methods. This behavior is what we would expect given that the CADD
annotation models only contain the most pathogenic sites, which are \emph{a
prior} very likely under the strong selection domain under which B' and classic
BGS theory agree.

The out-sample $R^2$

\begin{figure}[htbp] \centering
    \includegraphics[width=\textwidth]{figures/figure_2.pdf} 
    \caption{The $R^2$ estimates for a sparse (A) and full (B) models, for all 
    populations (colors) fit at the megabase-scale. Round points are our B' 
    method; diamonds are the classic B. For our method, we also calculate 
    leave-one-out out-sample $R^2$ (lighter filled points).}
  \label{fig:figure-2}
\end{figure}

\subsection*{Estimates of Negative Selection in Humans}

>>>>

\begin{figure}[htbp] \centering
    \includegraphics[width=\textwidth]{figures/figure_2.pdf} 
    \caption{The $R^2$ estimates for a sparse (A) and full (B) models, for all 
    populations (colors) fit at the megabase-scale. Round points are our B' 
    method and diamonds are the classic B. Lighter color round points are
    the out-sample $R_\text{LOO}^2$ estimates for our B' method, and arrows show the decline
    in goodness-of-fit due to in-sample overfitting (out-sample $R_\text{LOO}^2$ 
    were not calculated for classic B values due to computational costs). The 
    horizontal dashed lines are the $R_\text{drift}^2$ expected when the residual
    variance is given by the theoretic variance in coalescence times due to drift alone.}
  \label{fig:figure-2}
\end{figure}

Overall, we find the PhastCons Priority and CADD 6\% models fit about equally
as well (Figure \ref{fig:figure-2}A), explaining $R^2=68.45$\% and
$R^2=68.04$\% of the out-sample variance in Yoruba pairwise diversity at the
megabase-scale, respectively. In populations impacted by the out-of-Africa
bottleneck, the goodness-of-fit was lower across all models (e.g. 62.98\% and
59.98\% for CEU and CHB respectively in the PhastCons Priority model). The CADD
6\% model was previously found by \textcite{Murphy2022-sj} to be the
best-fitting model using classic BGS. Our PhastCons Priority model fits
marginally better than the CADD 6\% model consistently across all three
populations (Supplementary Table XXX). This is likely because the PhastCons
Priority model has more DFE parameters, and thus can fit finer-scale
differences in the DFE across its four feature types. By contrast, the CADD 6\%
track must fit patterns of diversity to conserved sites under a single DFE. The
CADD 8\% model fit slightly worse than the CADD 6\% across all populations, and
the Feature Priority model had the poorest-fit.

Since our method is built upon theory that fixes the weak selection problem of
classic BGS theory, it should in principle fit equally well when an annotation
model includes regions that are likely neutrally evolution. To test this, for
each annotation model (which are ``sparse") we fit a ``full" model that assigns
the remaining complement of the genome as a feature called ``other". This other
category is likely to predominantly neutrally evolving. Ideally, the
goodness-of-fit of model using our B' method should be invariant to whether an
annotation model is sparse or full. Indeed, we find this to be the case: both
in-sample $R^2$ and out-sample $R_\text{LOO}^2$ values are nearly identical
across full and sparse-track models (Figures \ref{fig:figure-2} A and B, round
points).

By contrast, full annotation models fit poorly when using classic BGS theory
(Figure \ref{fig:figure-2}B, diamond-shaped points) and lead to unreasonable
parameter estimates (discussed below). Additionally, when sparse annotation
models contain genomic features that are likely under weak constraint (such as
introns and UTRs), models fit worse under classic BGS theory than our B' method
(Figure \ref{fig:figure-2}A). However, among the CADD annotation models, the
goodness-of-fit is nearly identical between B' and classic BGS methods. This
behavior is what we would expect given that the CADD annotation models contain
only the most pathogenic sites, which are \emph{a prior} very likely under the
strong selection domain under which B' and classic BGS theory agree.

Along the chromosome, our approach models the observed megabase-scale diversity
in window $i$ as the predicted level under our negative selection model,
$B_i'$, plus some residual error $\varepsilon_i$, $\pi_i = B_i' +
\varepsilon_i$. The out-sample $R_\text{LOO}^2$ estimates the ratio of variance
explained by the negative selection model ($\var_i(\pi_i)$) to the total
variance in diversity along the genome ($\var_i(\pi_i) = \var_i(B_i) +
\var_i(\varepsilon_i) + 2\cov_i(B_i', \varepsilon_i)$, \cite{Murphy2022-sj}).
Since our $R_\text{LOO}^2$ estimates suggest our models fit left-out
chromosomes well even though our method assumes constant demography and
homogeneous mutation rates along the genome, we wondered what $R^2$ we could
expect under a simple theoretic model where $\cov_i(B_i', \varepsilon_i) = 0$
and $\var_i(\varepsilon_i)$ was determined by the variance in coalescence times
due to drift alone with an effective population size rescaled by $B_i'$. We
calculate this ``drift $R^2$" by calculating the theoretic coalescence time
variance in a megabase window with recombination and an effective population
size of $B_i' N_e$ where we use our best-fitting model's as a plug-in estimator
of $B_i'$. We note that this is a rough approximation, since this calculation
assumes constant demography and that the variance in coalescence times under
negative selection is likely lower than expected under drift alone. Still, we
find that our out-sample $R_\text{LOO}^2$ for Yoruba individuals
($R_\text{LOO}^2 \approx 68$\%) is close to the theoretic drift
$R_\text{drift}^2 \approx 67$\%. For bottlenecked out-of-Africa populations,
$R_\text{drift}^2 \approx 64$\% compared to the observed $R_\text{LOO}^2
\approx 62$\% for CEU and $R_\text{LOO}^2 \approx 59$\% for CHB. We note that
because bottlenecks increase the variance in coalescence times beyond the level
implied by the effective population size, the disagreement between observed
$R_\text{LOO}^2$ and the theoretic $R_\text{drift}^2$ is likely to shrink.
Overall, this suggests that negative selection models explain the vast majority
coalescence time variation at the megabase-scale that is capable of being
explained (i.e. that is not coalescence noise).

\subsection*{Estimates of Negative Selection Parameters in Humans}
